\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
updated ms figures

\subsection{Stastical models of vital rates}
The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level. Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level. We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below). We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrate level), and (3) a random effect of group (see Data set description) on the intercept. Vital rates for each species are modeled separately.

@@

\subsection{Computing}
\subsubsection{Model fitting}
All of our analyses (model fitting and simulating) were conducted in Program R. We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding the initial 20,000 iterations. We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
             caption="The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model. The values are model deviance estimated from the `coda' package in R."), 
      tabular.environment="tabularx", caption.placement="top",
      width="\\textwidth", include.rownames=FALSE)
@

% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
% # library(ggplot2)
% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
% #                     Contribution = c(growthSSRTable[,6],
% #                                      survivalSSRTable[,6],
% #                                      colonizationSSRTable[,6]))
% # ggplot(plotD, aes(x=Species, y=Contribution))+
% #   geom_bar(stat="identity")+
% #   facet_grid(.~Model)
% # @

>>>>

\subsection{Stastical models of vital rates}
The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level. Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level. We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below). We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept. Vital rates for each species are modeled separately.

@@

\subsection{Computing}
\subsubsection{Model fitting}
All of our analyses (model fitting and simulating) were conducted in Program R. We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in. We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

# print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
#              caption="The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model. The values are model deviance estimated from the `coda' package in R."), 
#       tabular.environment="tabularx", caption.placement="top",
#       width="\\textwidth", include.rownames=FALSE)
@

<<echo=FALSE, include=FALSE, fig=TRUE, label=ClimFig, width=5, height=3>>=
library(ggplot2)
devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
                       ConstantModel = rep(NA,3),
                       ClimateModel = rep(NA,3),
                       FullModel = rep(NA,3),
                       ClimateContribution = rep(NA,3))
devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
devM <- devGQ[,c(1,5)]
devM$Level <- rep("Individual", 3)
tmpD <- deviance[,c(1,5)]
tmpD$Level <- rep("Quadrat", 3)
colnames(tmpD)[2] <- "value"
colnames(tmpD)[1] <- "VitalRate"
colnames(devM)[2] <- "value"
plotD <- rbind(tmpD, devM)
ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
  geom_bar(stat="identity", position="dodge")+
  scale_fill_manual(values=c("steelblue", "darkorange"))+
  ylab("Contribution of Climate Effects")+
  xlab("Vital Rate")+
  theme_bw()
@

\begin{figure}[th!]
\begin{center}
\includegraphics{Fig-ClimFig}
\end{center}
\caption{The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization at the quadrat level, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model. The values are model deviance estimated from the `coda' package in R.}
\label{fig:ClimFig}
\end{figure}

### 2014-12-04 manuscript/ms_MicroMeso.tex
updated ms figures

\subsection{Including climate covariates (and avoiding model selection)}
For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We included climate covariates as additive effects. We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.

\subsection{Computing}
\subsubsection{Model fitting}
All of our analyses (model selection, fitting, and simulating) were conducted in R. We used the analytical Bayesian software INLA to fit candidate vital rate models at the individual and quadrat levels for computational efficiency. This is the case except for the recruitment model at the individual level which required a more complex hierarchical approach. In that case we used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters. After model selection, we used MCMC samplers, again in JAGS, to estimate the posterior distributions of model parameters for all vital rate models at the individual and quadrat levels. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 20,000 iterations, after discarding the initial 10,000 iterations. We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

% latex table generated in R 3.1.0 by xtable 1.7-3 package
% Tue Nov 25 11:51:40 2014
\begin{table}[ht]
\centering
\caption{The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model. The values are model deviance estimated from the `coda' package in R.} 
\begin{tabularx}{\textwidth}{llllX}
  \hline
Vital Rate & Constant model & Climate model & Full model & Contribution of climate covariates \\ 
  \hline
Growth & -5566.24 & -5735.32 & -6167.98 & 0.28 \\ 
  Survival & 171.55 & 170.37 &  & 0.01 \\ 
  Colonization & 203.98 & 193.65 &  & 0.05 \\ 
   \hline
\end{tabularx}
\end{table}
% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
% # library(ggplot2)
% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
% #                     Contribution = c(growthSSRTable[,6],
% #                                      survivalSSRTable[,6],
% #                                      colonizationSSRTable[,6]))
% # ggplot(plotD, aes(x=Species, y=Contribution))+
% #   geom_bar(stat="identity")+
% #   facet_grid(.~Model)
% # @

>>>>

\subsection{Including climate covariates (and avoiding model selection)}
For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We included climate covariates as additive effects. We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover. Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution. This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.

\subsection{Computing}
\subsubsection{Model fitting}
All of our analyses (model fitting and simulating) were conducted in Program R. We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in. We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

\begin{figure}[th!]
\begin{center}
\includegraphics{Fig-ClimFig}
\end{center}
\caption{The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model. The values are model deviance estimated from the `coda' package in R.}
\label{fig:ClimFig}
\end{figure}
